Gravitational field calculation in spherical coordinates using variable densities in depth

### 2018-12-28 manuscript/manuscript.tex
Edits to the introduction
I re-worded some sections to make things clearer and included more
citations to Lin and Denker. I re-wrote the part justifying why we need
a new method. The accuracy of the integration is probably not an issue
if you did a piecewise linear function (as our own results suggest).
Even using a high-degree polynomial wouldn't be too bad if the fit to
the density data is good. The main problem is that piecewise linear
raises the question of "what is the discretization interval" and "how
can we do this automatically"? We answer these with our algorithm. The
high-degree poly

To take into account the curvature of the Earth, many global forward modelling methods
are defined in geocentric spherical coordinates.
A common approach is to discretize the Earth into tesseroids \citep{Anderson1976},
i.e.~spherical prisms, which are defined by pairs of latitude, longitude, and radial
boundaries (see Fig.~\ref{fig:tesseroid}).
The gravitational fields generated by an arbitrary
tesseroid on any external point are given by volume
integrals that must be numerically approximated.
The literature offers two main approaches: one involves Taylor series expansion
\citep{Heck2007, Grombein2013} while the other makes use of Gauss-Legendre
Quadrature (GLQ)
\citep{Asgharzadeh2007, Wild-Pfeiffer2008, Li2011, Uieda2016}.
The Taylor series expansion is not well suited to develop an algorithm for
a density varying with depth according to an arbitrary continuous function.
Different series expansion terms would have to obtained for each density function
desired.
Conversely, an arbitrary continuous density function can be included in the GLQ without
any change to the integration method.

@@

The main challenge of the GLQ integration is loss of accuracy when the computation point
gets closer to the tesseroid \citep{Ku1977}.
\citet{Uieda2016} built on the three dimensional adaptive discretization algorithm of
\citet{Li2011} to automatically obtain integration results with 0.1\% accuracy.
The algorithm consist in recursively splitting the tesseroid into smaller ones when a
threshold is exceeded,
namely when the normalized distance to the computation point is greater than a
``distance-size ratio'' ($D$).
\citet{Uieda2016} have also obtained standard values of $D$
for the gravitational potential, its gradient and Marussi tensor components
by comparing the numerical model with the fields generated by a spherical shell.

Two recent papers present other approaches for homogeneous tesseroids gravitational
fields computation and incorporate methodologies for variable density tesseroid with
specific density profiles.
\citet{Fukushima2018} analytically integrates the volumetric integral for the
gravitational potential in the radial direction, obtaining a surface integral, which is
then numerically solved by applying a conditionally splitting and the double exponential
quadrature rule. The gradient of the potential and the Marussi tensor components are
then computed by finite differences. \citet{Fukushima2018} also generalizes his method
to an arbitrary polynomial density profile on the radial coordinate.
\citet{Lin2018} made a comparison of the different methodologies for homogeneous
tesseroid gravitational fields computation by applying different discretizations
algorithms to each one of them.
From this analysis they develop a combined method:
for computation points near the tesseroid they apply a GLQ numerical integration with
a two dimensional adaptive discretization algorithm, i.e.~only in the horizontal
dimension. But, if the computation point is farther than a certain truncation distance,
then a second order Taylor series approximation is applied along with the regular
subdivision developed by \citet{Grombein2013}.
They also introduce a variation of their combined method to compute the gravitational
fields generated by tesseroids with linear density on the radial coordinate.

In summary, the bibliography is offering different methodologies for computing
gravitational fields of homogeneous tesseroids or with certain specific density
functions in depth.
Although any continuous density profile could be approximated by one of these specific
density functions, one should acknowledge the error that such approximation would
propagate into the gravitational field in order to guarantee certain degree of accuracy.

We present a new algorithm for computing the gravitational fields generated by any
tesseroid with any arbitrary continuous density function on any external point.
It is based on the three dimensional GLQ approximation, a two dimensional version of the
adaptive discretization of \citet{Uieda2016} and a new density-based discretization
algorithm that we introduce in this work.
In order to ensure the accuracy of the numerical approximation we have
determined optimal values for the controlling parameters by
comparing the numerical approximation with analytical solutions for
spherical shells.
Finally, we applied the methodology to model the Neuqu\'en basin, Argentina, using
tesseroids with linear and exponentially increasing density with depth.

>>>>

To take into account the curvature of the Earth, many global forward modelling methods
are defined in geocentric spherical coordinates.
A common approach is to discretize the Earth into tesseroids \citep{Anderson1976},
i.e.~spherical prisms, which are defined by pairs of latitude, longitude, and radial
boundaries (Fig.~\ref{fig:tesseroid}).
The gravitational fields generated by a
tesseroid on any external point are given by volume
integrals that must be numerically approximated.
The literature offers two main approaches: one involves Taylor series expansion
\citep{Heck2007, Grombein2013} while the other makes use of Gauss-Legendre
Quadrature (GLQ)
\citep{Asgharzadeh2007, Wild-Pfeiffer2008, Li2011, Uieda2016, Lin2018}.
The Taylor series expansion is not well suited to develop an algorithm for
a density varying with depth according to an arbitrary continuous function.
Different series expansion terms would have to obtained for each density function
desired.
Conversely, an arbitrary continuous density function can be included in the GLQ without
any change to the integration method.
Thus, we will focus only on GLQ-based methods henceforth.

@@

The main challenge of the GLQ integration is the loss of accuracy that occurs when the
computation point approaches the tesseroid \citep{Ku1977}.
\citet{Uieda2016} built on the three dimensional adaptive discretization algorithm of
\citet{Li2011} to automatically obtain integration results with 0.1\% accuracy.
The algorithm recursively divides the tesseroid into smaller ones when a
threshold is exceeded,
namely when the normalized distance to the computation point is greater than a
``distance-size ratio'' parameter ($D$).
\citet{Uieda2016} have also obtained standard values of $D$
for the gravitational potential, its gradient and Marussi tensor components
by comparing the numerical integration results with the fields generated by a spherical
shell.

Two recent publications present alternative approaches for calculating the gravitational
fields of homogeneous tesseroids and incorporate methodologies for tesseroids with
variable densities in depth.
\citet{Fukushima2018} analytically integrates the volumetric integral for the
gravitational potential in the radial direction, obtaining a surface integral, which is
then numerically solved by conditionally splitting the tesseroid and applying the double
exponential quadrature rule.
The gradient of the potential and the Marussi tensor components are computed by finite
differences.
\citet{Fukushima2018} also generalized their method to tesseroids with a radial
polynomial density function of arbitrary degree.
\citet{Lin2018} compared the different integration and discretization methodologies for
homogeneous tesseroids.
From this analysis they developed a combined method:
for computation points near the tesseroid, they use a GLQ integration with an adaptive
discretization based on \citet{Uieda2016} but only applied to the horizontal dimensions.
If the computation point is farther than a certain truncation distance,
a second order Taylor series approximation is applied instead along with the regular
subdivision developed by \citet{Grombein2013}.
\citet{Lin2018} also introduced a variation of their combined method to compute the
gravitational fields generated by tesseroids with a linearly varying density in the
radial dimension.

Both the \citet{Lin2018} and the \citet{Fukushima2018} studies limit the radial density
variation to polynomial functions.
While most continuous and smooth functions can be approximated by piecewise linear
functions, the choice of a discretization interval is neither straight forward nor
automatic.
Furthermore, it is well known that the use of high-degree polynomials to approximate a
highly variable function produces unstable results when extrapolating beyond the data
domain.
These shortcomings could make piecewise linear or high-degree polynomial density
functions unwieldy for non-linear gravity inversions if the density function in depth
is highly variable.

We present a new algorithm for computing the gravitational fields generated by a
tesseroid with an arbitrary continuous density function on an external point.
It is based on the three dimensional GLQ integration, a two dimensional version of the
adaptive discretization of \citet{Uieda2016} (following \citet{Lin2018}),
and a new density-based radial discretization algorithm.
To ensure the accuracy of the numerical approximation, we empirically determine
optimal values for the controlling parameters by comparing the numerical results with
analytical solutions for spherical shells.
Finally, we apply the methodology to model the Neuqu\'en basin, Argentina, using
tesseroids with linear and exponentially increasing density with depth.
